Sampling Strategies in Decisions from Experience

### 2022-04-22 manuscript.Rmd
Minor reviews

shorttitle        : "Sampling strategies"

@@

authornote: |
 This manuscript is a dynamic document which can be reproduced using the materials and instructions provided on the `GitHub` repository: https://github.com/linushof/sampling-in-dfe. The current version of the manuscript, created from the commit with the hash `r repro::current_hash()`, is prepared for submission as a master's thesis in fulfillment of the requirements for the degree Master of Science (M.Sc.) at Heidelberg University, Faculty of Behavioural and Cultural Studies, Institute of Psychology. Supervisors:

@@

abstract: "In decisions from experience, systematic choices between prospects are based on the assessment of differences in latent properties. Yet, several properties can be used to distinguish among the prospects. To this end, this paper presents a sampling and accumulation model that allows sampling  strategies—i.e., the probability of sampling outcomes from different prospects in direct succession (switching probability)—to alter which differences in the latent properties evidence is accumulated for. Specifically, the proposed model sequentially samples outcomes from two prospects and accumulates the results obtained from a sequence of ordinal comparisons between sample means, with switching probabilities determining the size of the samples underlying each comparison and thresholds determining how many comparisons a prospect must win in order to be chosen. A computational implementation of the model is used to simulate the respective information-processing chains upfront choices between a safe and a two-outcome prospect, while varying switching probabilities and thresholds. The analysis of the simulation data shows that, depending on whether the accumulation process unfolds over mean comparisons based on small or large samples, evidence is accumulated for differences either in the probability to produce a higher outcome most of the time or in the expected value. The respective choice patterns reflect an as-if underweighting of rare outcomes or expected value maximization, both of which become more robust with increasing thresholds. The analysis moreover shows how the distinct choice patterns resulting from differences in the information-processing chains translate to the parameters of cumulative prospect theory’s value and weighting function."

@@

bibliography      : references_sampling-in-dfe.bib
csl               : apa.csl

@@

header-includes:
   - \usepackage[capposition=top]{floatrow}
---

@@

<!-- Introduction -->

@@

Now, let a *prospect* be a choice alternative that possesses only two kinds of fundamental properties: the possible outcomes of the alternative (e.g., gains or losses of some amount) and the probabilities with which these outcomes occur following the choice of the alternative.
Importantly, the information that the inputs provide about these properties can come in different forms.
Specifically, in *decisions from description* (DFD), the inputs take the form of explicit and complete descriptions of all outcomes and probabilities.
Yet, in *decisions from experience* [DFE, @hertwigDecisionsExperienceEffect2004], these properties are latent and therefore not known with certainty, but they must be inferred from the relative frequencies with which the outcomes occurred in the past, e.g., when the prospect was chosen in similar past decisions.
The inputs to DFE therefore take the form of sampled outcomes.
Although these distinct input forms may in principle carry the same information about a given set of prospects, behavioral decision research has so far produced a large body of papers indicating that---taking a quite general perspective---there are robust differences between the choices that are made in DFD and DFE, leading to the notion of the *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009].
In one reading of the gap,^[<!--Open footnote-->See @hertwigConstructbehaviorGapDescription2018, for a discussion on the interpretation of the gap.<!--Close footnote-->]
choice patterns in DFD deviate from the solutions of expected value (EV) maximization, as if small-probability outcomes are given more weight than would be warranted by their objective probabilities;
in turn, in DFE, choice patterns deviate from the solutions of EV maximization, as if the same small-probability outcomes are given less weight than would be objectively warranted [e.g., @barronSmallFeedbackbasedDecisions2003; @hertwigDecisionsExperienceEffect2004; @weberPredictingRiskSensitivity2004; @erevAnomaliesForecastsDescriptive2017; @regenwetterConstructbehaviorGapBehavioral2017; see @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-analytic review].
Hereafter, these choice patterns are referred to as the *as-if overweighting* and *as-if underweighting of rare outcomes*, respectively.

@@

To substantiate the discussion, the model is used to simulate the sampling and accumulation process for choices between a safe and a two-outcome prospect, while systematically combining different switching probabilities and thresholds.
Considering choice rates and parameter estimates of cumulative prospect theory [@tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993], the computational analysis shows that the choices converge to the solutions of EV maximization for low switching probabilities, but take the form of an as-if underweighting of rare outcomes for high switching probabilities.
Importantly, the analysis also shows that these distinct choice patterns become more robust with increasing thresholds.
Before presenting the model and the computational analysis, the following section provides a formal definition of prospects.

@@

Since there is an uncountable number of choices people confront in the natural world, behavioral decision research routinely abstracts from choices between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two prospects, it studies a case which is counterfactual in that it omits the many particularities of each choice situation, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess.^[<!--Open footnote-->See @knightRiskUncertaintyProfit1921, Chapter 1, for a timeless discussion on the reasonableness of such abstractions.<!--Close footnote-->]
These fundamental properties are considered the possible outcomes of a choice alternative and the probabilities with which these outcomes occur following the choice of the alternative.
Roughly speaking, then, all choice alternatives that can be fully described by their outcome-probability pairs are prospects.
As such, prospects can be understood as the distribution of a discrete random variable.
More specifically, following @kolmogorovFoundationsTheoryProbability1950 and adopting the notation from @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, let the tuple $(\Omega, \mathcal{F}, P)$ be a probability space.
Accordingly, $\Omega$ is a finite sample space with each element $\omega \in \Omega$ denoting a possible consequence of choosing the respective prospect.
$\mathcal{F}$ is the power set $\mathcal{P}(\Omega)$ with each element $A \in \mathcal{F}$ being a subset of $\Omega$, also denoted a random event.
$P$ is the probability measure
$$
P: \mathcal{F} \to [0,1]
\tag{1}
$$
assigning the elements of $\mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.
Now, the random variable is a function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{2}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ denoting a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the pushforward measure
$$
P' : \mathcal{F'} \to [0,1]
\tag{3}
$$
with 
$$
P'(A' \in \mathcal{F'}) := P(\{\omega: X(\omega) = x \in A'\})
\; .
\tag{4}
$$
Note that the measure in (3) and (4), respectively, is a random variable's distribution and therefore a prospect. 
Accordingly, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
\{P'^1, ... , P'^k\}
\; ,
\tag{5}
$$
where each element $P'$ can itself be considered a finite set of outcome-probability pairs $\{(x_i, p_X(x_i)\}_{i \in \mathbb{N}}$.

@@

Yet, there are several distinct properties that can be used to assess differences between the prospects, among them the EV, for which the summary decision strategy is indicative, and the probability to produce a higher outcome most of the time, for which the round-wise decision strategy is indicative [see @hillsInformationSearchDecisions2010].
Respectively, in the following, both decision strategies are implemented into a common sampling and accumulation model to demonstrate how the mind may rely on different sampling strategies to alter the degree to which evidence is accumulated for differences in either one of these properties.
More specifically, the proposed model accumulates the outcomes obtained from a sequence of comparisons between sample means, with the size of the samples underlying each mean comparison tending to be large (small) for low (high) switching probabilities.
Accordingly, whereas the decision variables of the model capture how many comparisons a prospect has won, the unfolding accumulation process more closely resembles the summary strategy for low switching probabilities and the round-wise strategy for high switching probabilities.

@@

To substantiate and refine the predictions of the model, a computational implementation was used to simulate the sampling and accumulation process for a set of choices between a safe prospect and a two-outcome (risky) prospect.
The computational model was written in `R` and is embodied in the `GitHub` repository accompanying this paper.
The simulation was conducted for varying combinations of $\psi$ and $\theta$.
For each of these combinations, the simulated data was modeled with a stochastic version of CPT, using Bayesian parameter estimation.
The simulated choices and estimated CPT parameters can be roughly understood as theory-implied data and data models, which remain to be tested against empirical data [see, e.g., @guestHowComputationalModeling2021; @haslbeckModelingPsychopathologyData2021].

@@

The choice data for each parameter combination was separately modeled in CPT, using the value function of @tverskyAdvancesProspectTheory1992 and the weighting function of @goldsteinExpressionTheoryPreference1987.
To demonstrate that the distortions from linear probability weighting are not just due to differences between the latent objective probabilities and the relative frequencies with which the outcomes were actually sampled over the course of a sampling and accumulation process, the latter were supplied as probability information to the weighting function.
To accommodate for possible random choice behavior that may be exerted by a parameter combination, the CPT model was amended with the logit choice rule
$$
p(P'^{safe}) = \frac{1} {1 + e^{-\rho(V_{safe}^{\frac{1}{\alpha}}-V_{risky}^{\frac{1}{\alpha}})}}
\; ,
\tag{16}
$$
where $p(P'^{safe})$ is the probability that the safe prospect is chosen over the risky prospect, $V_{safe}$ and $V_{risky}$ are the valuations of the safe and risky prospect that result from the parameter estimates of the value and weighting function, and $\rho \geq 0$ is a sensitivity parameter.
$\rho$ governs how strongly the choice between prospects depends on the differences in the valuations of the prospects, i.e., the core CPT model.
Specifically, $\rho = 0$ implies no sensitivity, and thus a random choice.
In turn, for increasing values of $\rho$, the probability of choosing the prospect with the higher valuation increases.
Note further that the valuation of the prospects were rescaled by $\frac{1}{\alpha}$, as proposed by @stewartPsychologicalParametersHave2018, to guard against possible parameter interdependencies between $\alpha$ and $\rho$ [@krefeld-schwalbStructuralParameterInterdependencies2021; @scheibehenneUsingBayesianHierarchical2015].
In sum, then, the model has four free parameters: $\alpha, \gamma, \delta, \rho$.

@@

```{r data, include=FALSE}
#load choice data
cols_choices <- list(.default = col_double(),
                     boundary = col_factor(),
                     gamble = col_factor(),
                     rare = col_factor(),
                     agent = col_factor(),
                     choice = col_factor())
choices <- read_csv("data/choices/choices.csv", col_types = cols_choices)
choices <- choices %>% 
  filter(boundary == "absolute")

#load CPT estimates
cols_cpt <- list(.default = col_double(),
                 boundary = col_factor(),
                 a = col_factor(),
                 parameter = col_factor())
cpt <- read_csv("data/estimates/estimates_cpt.csv", col_types = cols_cpt)
cpt <- cpt %>% 
  filter(boundary == "absolute")
```

To estimate the posterior distributions of the free parameters, Markov Chain Monte Carlo sampling as implemented in `JAGS` was used, where 20 chains of 40,000 samples each were run after a warm-up period of 1000 samples.
To reduce potential autocorrelation during the sampling process, only every 20th sample was kept (thinning), leaving a total of 40,000 samples across chains.
The minimum effective sample size was `r min(cpt$n.eff)`.
The potential scale reduction factor $\hat{R}$ [@gelmanInferenceIterativeSimulation1992] was $\leq$ `r round(max(cpt$Rhat), 4)` for all models and parameters, indicating good convergence.

>>>>

shorttitle        : |
  Sampling strategies

@@

authornote: |
 This manuscript is a dynamic document which can be reproduced using the materials and instructions provided on the GitHub repository: https://github.com/linushof/sampling-in-dfe. The current version of the manuscript, created from the commit with the hash `r repro::current_hash()`, is prepared for submission as a master's thesis in fulfillment of the requirements for the degree Master of Science (M.Sc.) at Heidelberg University, Faculty of Behavioural and Cultural Studies, Institute of Psychology. Supervisors:

@@

abstract: |
  In decisions from experience, the probability to produce a higher outcome most of the time and the expected value are two latent properties that can be used to distinguish among the prospects of choice. This paper sketches a sampling and accumulation model that allows sampling strategies—i.e., the probability of sampling outcomes from different prospects in direct succession (switching probability)—to alter the degree to which evidence is accumulated for either one of these properties. A computational implementation of the model is used to simulate the respective sampling and accumulation process upfront choices between a safe and a two-outcome prospect, while varying switching probabilities and evidence thresholds. The simulation data shows that, depending on which latent property evidence is accumulated for and to which extent, respectively, choice patterns reflect various degrees of an as-if underweighting of rare outcomes or expected value maximization. Both choice patterns become more robust the more evidence is sampled and translate to characteristic parameter estimates for cumulative prospect theory’s value and weighting function.

@@

bibliography      : references.bib
csl               : apa.csl

@@

header-includes:
   - |
    \usepackage[capposition=top]{floatrow}
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
---

@@

@@

Now, let a *prospect* be a choice alternative that possesses only two kinds of fundamental properties: the possible outcomes of the alternative (e.g., gains or losses of some amount) and the probabilities with which these outcomes occur following the choice of the alternative.
Importantly, the information that the inputs provide about these properties can come in different forms.
Specifically, in *decisions from description* (DFD), the inputs take the form of explicit and complete descriptions of all outcomes and probabilities.
Yet, in *decisions from experience* [DFE, @hertwigDecisionsExperienceEffect2004], these properties are latent and therefore not known with certainty, but they must be inferred from the relative frequencies with which the outcomes occurred in the past, e.g., when the prospect was chosen in similar past decisions.
The inputs to DFE therefore take the form of sampled outcomes.
Although these distinct input forms may in principle carry the same information about a given set of prospects, behavioral decision research has so far produced a large body of papers indicating that---taking a quite general perspective---there are robust differences between the choices that are made in DFD and DFE, leading to the notion of the *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009].
In one reading of the gap,^[<!--Open footnote-->See @hertwigConstructbehaviorGapDescription2018, for a discussion on the interpretation of the gap.<!--Close footnote-->]
choice patterns in DFD deviate from the solutions of expected value (EV) maximization, as if small-probability outcomes are given more weight than would be warranted by their objective probabilities (*as-if overweighting of rare outcomes*);
in turn, in DFE, choice patterns deviate from the solutions of EV maximization, as if the same small-probability outcomes are given less weight than would be objectively warranted [*as-if underweighting of rare outcomes*, e.g., @barronSmallFeedbackbasedDecisions2003; @hertwigDecisionsExperienceEffect2004; @weberPredictingRiskSensitivity2004; @erevAnomaliesForecastsDescriptive2017; @regenwetterConstructbehaviorGapBehavioral2017; see @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-analytic review].

@@

To substantiate the discussion, the model is used to simulate the sampling and accumulation process for choices between a safe and a two-outcome prospect, while systematically combining different switching probabilities and thresholds.
Considering choice rates and parameter estimates of cumulative prospect theory [@tverskyAdvancesProspectTheory1992], the computational analysis shows that the choices converge to the solutions of EV maximization for low switching probabilities, but take the form of an as-if underweighting of rare outcomes for high switching probabilities.
Importantly, the analysis also shows that these distinct choice patterns become more robust with increasing thresholds.
Before presenting the model and the computational analysis, the following section provides a formal definition of prospects.

@@

Since there is an uncountable number of choices people confront in the natural world, behavioral decision research routinely abstracts from choices between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two prospects, it studies a case which is counterfactual in that it omits the many particularities of each choice situation, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess.^[<!--Open footnote-->See @knightRiskUncertaintyProfit1921, Chapter 1, for a discussion on the reasonableness of such abstractions.<!--Close footnote-->]
These fundamental properties are considered the possible outcomes of a choice alternative and the probabilities with which these outcomes occur following the choice of the alternative.
Roughly speaking, then, all choice alternatives that can be fully described by their outcome-probability pairs are prospects.
As such, prospects can be understood as the distribution of a discrete random variable.
More specifically, following @kolmogorovFoundationsTheoryProbability1950 and adopting the notation from @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, let the tuple $(\Omega, \mathcal{F}, P)$ be a probability space.
Accordingly, $\Omega$ is a finite sample space with each element $\omega \in \Omega$ denoting a possible consequence of choosing the respective prospect.
$\mathcal{F}$ is the power set $\mathcal{P}(\Omega)$ with each element $A \in \mathcal{F}$ being a subset of $\Omega$, also denoted a random event.
$P$ is the probability measure
$$
P: \mathcal{F} \to [0,1]
\tag{1}
$$
assigning the elements of $\mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.
Now, the random variable is a function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{2}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ denoting a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the pushforward measure
$$
P' : \mathcal{F'} \to [0,1]
\tag{3}
$$
with 
$$
P'(A' \in \mathcal{F'}) := P(\{\omega: X(\omega) = x \in A'\})
\; .
\tag{4}
$$
Note that the measure in (3) and (4), respectively, is a random variable's distribution and therefore a prospect. 
Accordingly, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
\{P'^1, ... , P'^k\}
\; ,
\tag{5}
$$
where each element $P'$ can itself be considered a finite set of outcome-probability pairs $\{(x_i, p_X(x_i)\}_{i \in \mathbb{N}}$.

@@

Yet, there are several distinct properties that can be used to assess differences between the prospects, among them the EV, for which the summary decision strategy is indicative, and the probability to produce a higher outcome most of the time, for which the round-wise decision strategy is indicative [@hillsInformationSearchDecisions2010; see also @wulffHowShortLongrun2015].
Respectively, in the following, both decision strategies are implemented into a common sampling and accumulation model to demonstrate how the mind may rely on different sampling strategies to alter the degree to which evidence is accumulated for differences in either one of these properties.
More specifically, the proposed model accumulates the outcomes obtained from a sequence of comparisons between sample means, with the size of the samples underlying each mean comparison tending to be large (small) for low (high) switching probabilities.
Accordingly, whereas the decision variables of the model capture how many comparisons a prospect has won, the unfolding accumulation process more closely resembles the summary strategy for low switching probabilities and the round-wise strategy for high switching probabilities.

@@

To substantiate and refine the predictions of the model, a computational implementation was used to simulate the sampling and accumulation process for a set of choices between a safe prospect and a two-outcome (risky) prospect.
The computational model was written in R and is embodied in the GitHub repository accompanying this paper.
The simulation was conducted for varying combinations of $\psi$ and $\theta$.
For each of these combinations, the simulated data was modeled with a stochastic version of CPT, using Bayesian parameter estimation.
The simulated choices and estimated CPT parameters can be roughly understood as theory-implied data and data models, which remain to be tested against empirical data [see, e.g., @guestHowComputationalModeling2021; @haslbeckModelingPsychopathologyData2021].

@@

The choice data for each parameter combination was separately modeled in CPT, using the value function of @tverskyAdvancesProspectTheory1992 and the weighting function of @goldsteinExpressionTheoryPreference1987.
To demonstrate that the distortions from linear probability weighting are not just due to differences between the latent objective probabilities and the relative frequencies with which the outcomes were actually sampled over the course of a sampling and accumulation process, the latter were supplied as probability information to the weighting function.
To accommodate for possible random choice behavior that may be exerted by a parameter combination, the CPT model was amended with the logit choice rule
$$
p(P'^{safe}) = \frac{1} {1 + e^{-\rho(V_{safe}^{\frac{1}{\alpha}}-V_{risky}^{\frac{1}{\alpha}})}}
\; ,
\tag{16}
$$
where $p(P'^{safe})$ is the probability that the safe prospect is chosen over the risky prospect, $V_{safe}$ and $V_{risky}$ are the valuations of the safe and risky prospect that result from the parameter estimates of the value and weighting function, and $\rho \geq 0$ is a sensitivity parameter.
$\rho$ governs how strongly the choice between prospects depends on the differences in the valuations of the prospects, i.e., the core CPT model.
Specifically, $\rho = 0$ implies no sensitivity, and thus a random choice.
In turn, for increasing values of $\rho$, the probability of choosing the prospect with the higher valuation increases.
Note further that the valuation of the prospects were rescaled by $\frac{1}{\alpha}$, as proposed by @stewartPsychologicalParametersHave2018, to guard against possible parameter interdependencies between $\alpha$ and $\rho$ [@krefeld-schwalbStructuralParameterInterdependencies2022; @scheibehenneUsingBayesianHierarchical2015].
In sum, then, the model has four free parameters: $\alpha, \gamma, \delta, \rho$.

@@

```{r data, include=FALSE}
#load choice data
cols_choices <- list(.default = col_double(),
                     boundary = col_factor(),
                     gamble = col_factor(),
                     rare = col_factor(),
                     agent = col_factor(),
                     choice = col_factor())
choices <- read_csv("data/simulation_summary.csv", col_types = cols_choices)
choices <- choices %>% 
  filter(boundary == "absolute")

#load CPT estimates
cols_cpt <- list(.default = col_double(),
                 boundary = col_factor(),
                 a = col_factor(),
                 parameter = col_factor())
cpt <- read_csv("data/cpt_parameters.csv", col_types = cols_cpt)
cpt <- cpt %>% 
  filter(boundary == "absolute")
```

To estimate the posterior distributions of the free parameters, Markov Chain Monte Carlo sampling as implemented in JAGS was used, where 20 chains of 40,000 samples each were run after a warm-up period of 1000 samples.
To reduce potential autocorrelation during the sampling process, only every 20th sample was kept (thinning), leaving a total of 40,000 samples across chains.
The minimum effective sample size was `r min(cpt$n.eff)`.
The potential scale reduction factor $\hat{R}$ [@gelmanInferenceIterativeSimulation1992] was $\leq$ `r round(max(cpt$Rhat), 4)` for all models and parameters, indicating good convergence.
